---
title: "bcbioR quick start"
output: rmarkdown::html_vignette
vignette: >

  %\VignetteIndexEntry{bcbioR quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bcbioR)
library(ggplot2)
library(ggprism)

colors=cb_friendly_cols(1:15)
ggplot2::theme_set(theme_prism(base_size = 14))
```

# Color blind friendly palette

Compatible with `ggplot`.

```{r, fig.width=7}
set.seed(596)
dsamp <- diamonds[sample(nrow(diamonds), 1000), ]
ggplot(dsamp, aes(carat, price)) +
  geom_point(aes(colour = clarity)) +
  scale_color_cb_friendly()
```


And get the colors directly:

```{r}
cb_friendly_cols(1:16)
```
This is the full palette:

```{r fig.width=9, fig.height=6}
library(hues)
swatch(cb_friendly_cols(1:16))
```

# Set projects

HCBC uses a structured based directory to organize projects. You can set up this by using:

```{r}
tmp_dir=withr::local_tempdir()
bcbio_templates(type="base", outpath=tmp_dir)
fs::dir_ls(tmp_dir, recurse=TRUE)
```


We support multiple analyses type:

- RNAseq
- scRNAseq
- ChipPseq

To get the example code for any of them you can use a similar command:

```{r}
analysis_tmp=fs::path_join(c(tmp_dir, "reports"))
bcbio_templates(type="rnaseq", outpath=analysis_tmp)
fs::dir_ls(analysis_tmp, recurse=TRUE)
```

Use `scrnaseq`, `teaseq` or `cosmx` to get those other templates.
