---
title: "Quality Control"
author: "Harvard Chan Bioinformatics Core"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
params:
  # Fill this file with the right paths to nfcore output
  # Put hg38, mm10, mm39, or other
  # params_file: params_qc-example.R # example data
  params_file: params_qc-example.R
  genome: hg38
  single_end: false
  factor_of_interest: sample_type
  project_file: ../information.R
  functions_file: ../00_libs/load_data.R
---

Template developed with materials from https://hbctraining.github.io/main/.

```{r, cache = FALSE, message = FALSE, warning=FALSE}
# This set up the working directory to this file so all files can be found
library(rstudioapi)
setwd(fs::path_dir(getSourceEditorContext()$path))
# NOTE: This code will check version, this is our recommendation, it may work
#.      other versions
stopifnot(R.version$major>= 4) # requires R4
if (compareVersion(R.version$minor,"3.1")<0) warning("We recommend >= R4.3.1") 
stopifnot(compareVersion(as.character(BiocManager::version()), "3.18")>=0)
```

This code is in this ![](https://img.shields.io/badge/status-stable-green) revision.

```{r source_params, cache = FALSE, message = FALSE, warning=FALSE}
# 1. set up factor_of_interest parameter from parameter above or manually
#    this is used to color plots, it needs to be part of the metadata
factor_of_interest=params$factor_of_interest
genome=params$genome
<<<<<<< HEAD
singleend=params$singleend
=======
single_end=params$single_end
>>>>>>> c37fcd42e9e4ba276293a9bc591e6298affd14e5
# 2. Set input files in this file
source(params$params_file)
# 3. If you set up this file, project information will be printed below and
#.   it can be reused for other Rmd files.
source(params$project_file)
# 4. Load custom functions to load data from coldata/metrics/counts
source(params$functions_file)
```

# Overview

-   Project: `r project`
-   PI: `r PI`
-   Analyst: `r analyst`
-   Experiment: `r experiment`


```{r load_libraries, cache = FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(knitr)
library(rtracklayer)
library(DESeq2)
library(DEGreport)
library(ggrepel)
# library(RColorBrewer)
library(DT)
library(pheatmap)
library(RColorBrewer)
library(ggprism)
library(grafify)
ggplot2::theme_set(theme_prism(base_size = 12))
# https://grafify-vignettes.netlify.app/colour_palettes.html
# NOTE change colors here if you wish
scale_colour_discrete <- function(...)
  scale_colour_manual(..., 
                      values = as.vector(grafify:::graf_palettes[["kelly"]]))
scale_fill_discrete <- function(...)
  scale_fill_manual(..., 
                      values = as.vector(grafify:::graf_palettes[["kelly"]]))

opts_chunk[["set"]](
    cache = FALSE,
    cache.lazy = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE,
    fig.height = 4)
```


```{r sanitize-datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```


# Samples and metadata

```{r load_data, message=F, warning=F}
# This code will load from bcbio or nf-core folder
# NOTE make sure to set numerator and denominator
coldata <- load_coldata(coldata_fn)
coldata$sample=row.names(coldata)
 
counts <- load_counts(counts_fn)
counts <- counts[,colnames(counts) %in% coldata$sample]

metrics <- load_metrics(se_object, multiqc_data_dir,
                        gtf_fn, counts, single_end) %>% 
  left_join(coldata, by = c('sample')) %>% 
  as.data.frame()
metrics <- subset(metrics, metrics$sample %in% coldata$sample)
# TODO: change order as needed
order <- unique(metrics[["sample"]])
rownames(metrics) <- metrics$sample
# if the names don't match in order or string check files names and coldata information
counts = counts[,rownames(metrics)]
coldata = coldata[rownames(metrics),]
stopifnot(all(names(counts) == rownames(metrics)))
```

```{r load_metadata}
meta_df=coldata
ggplot(meta_df, aes(.data[[factor_of_interest]],
                    fill = .data[[factor_of_interest]])) +
  geom_bar() + ylab("") + xlab("") + ylab("# of samples") +
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5),
        legend.position = "none")
```


```{r show_metadata}
meta_sm <- meta_df %>%
  as.data.frame() 

meta_sm %>% sanitize_datatable()
```

# Read metrics {.tabset}

## Total reads

Here, we want to see consistency and a minimum of 20 million reads (the grey line).

```{r plot_total_reads}
metrics %>%
  ggplot(aes(x = factor(sample, level = order),
             y = total_reads/1000000, 
             fill = .data[[factor_of_interest]])) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_y_continuous(name = "Millions of reads") +
  scale_x_discrete(limits = rev) +
  xlab("") + 
  ggtitle("Total reads") +
  geom_hline(yintercept=20, color = "grey", linewidth=2)

metrics %>%
    ggplot(aes(x = .data[[factor_of_interest]],
               y = total_reads, 
               color = .data[[factor_of_interest]])) +
    geom_point(alpha = 0.5, size=4) +
    coord_flip() +
    scale_y_continuous(name = "million reads") +
    ggtitle("Total reads") 
```

```{r calc_min_max_pct_mapped}
#get min percent mapped reads for reference
min_pct_mapped <- round(min(metrics$mapped_reads/metrics$total_reads)*100,1)
max_pct_mapped <- round(max(metrics$mapped_reads/metrics$total_reads)*100,1)
```

## Mapping rate

The genomic mapping rate represents the percentage of reads mapping to the reference genome. We want to see consistent mapping rates between samples and over 70% mapping (the grey line). These samples have mapping rates: `r min_pct_mapped` - `r max_pct_mapped`%.

```{r plot_mapping_rate}
metrics$mapped_reads_pct <- round(metrics$mapped_reads/metrics$total_reads*100,1)
metrics %>%
    ggplot(aes(x = factor(sample, level = order), 
               y = mapped_reads_pct, 
               color = .data[[factor_of_interest]])) +
    geom_point(alpha = 0.5, size=4) +
    coord_flip() +
    ylab("Mapped Reads %") +
    scale_x_discrete(limits = rev) +
    ylim(0, 100) +
  ggtitle("Mapping rate") + xlab("") +
  geom_hline(yintercept=70, color = "grey", linewidth=2)
```


## Number of genes detected

The number of genes represented in every sample is expected to be consistent and over 20K (grey line).

```{r calc_genes_detected}
genes_detected <- colSums(counts > 0) %>% enframe()
sample_names <- metrics[,c("sample"), drop=F]
genes_detected <- left_join(genes_detected, sample_names, by = c("name" = "sample"))
genes_detected <- genes_detected %>% group_by(name)
genes_detected <- summarise(genes_detected, 
                             n_genes = max(value))
                      
metrics <- metrics %>%
  left_join(genes_detected, by = c("sample" = "name"))

```


```{r plot_genes_detected}
ggplot(metrics,aes(x = factor(sample, level = order),
           y = n_genes, fill = .data[[factor_of_interest]])) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  ggtitle("Number of genes") +
  ylab("Number of genes") +
  xlab("") +
  geom_hline(yintercept=20000, color = "grey", linewidth=2)

metrics %>%
  ggplot(aes(x = .data[[factor_of_interest]],
             y = n_genes, 
             color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  scale_y_continuous(name = "million reads") +
  xlab("") + 
  ggtitle("Number of Genes") 

```

## Gene detection saturation

This plot shows how complex the samples are. We expect samples with more reads to detect more genes. 

```{r plot_gene_saturation}
metrics %>% 
  ggplot(aes(x = total_reads/1000000, 
             y = n_genes,
             color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  scale_x_log10() +
  ggtitle("Gene saturation") +
  ylab("Number of genes") +
  xlab("Millions of reads")
```

## Exonic mapping rate

Here we are looking for consistency, and exonic mapping rates around or above 70% (grey line). 

```{r plot_exonic_mapping_rate}
metrics %>%
  ggplot(aes(x = factor(sample, level = order),
             y = exonic_rate * 100, 
             color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  ylab("Exonic rate %") + 
  ggtitle("Exonic mapping rate") + 
  scale_x_discrete(limits = rev) +
  coord_flip()  +
  xlab("") +
  ylim(c(0,100)) +
  geom_hline(yintercept=70, color = "grey", linewidth=2) 
```

## Intronic mapping rate

Here, we expect a low intronic mapping rate (≤ 15% - 20%). The grey line indicates 20%.

```{r plot_intronic_mapping_rate}
metrics %>%
    ggplot(aes(x = factor(sample, level = order),
               y = intronic_rate * 100, 
               color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  ylab("Intronic rate %") +
  ggtitle("Intronic mapping rate") + 
  scale_x_discrete(limits = rev) +
  coord_flip()  +
  xlab("") +
  ylim(c(0,100)) +
  geom_hline(yintercept=20, color = "grey", linewidth=2) 
```

## Intergenic mapping rate

Here, we expect a low intergenic mapping rate, which is true for all samples. The grey line indicates 15%

```{r plot_intergenic_mapping_rate}
metrics %>%
    ggplot(aes(x = factor(sample, level = order),
               y = intergenic_rate * 100, 
               color = .data[[factor_of_interest]])) +
    geom_point(alpha = 0.5, size=4) +
    ylab("Intergenic rate %") +
    ggtitle("Intergenic mapping rate") + 
    coord_flip()  + xlab("") +
    scale_x_discrete(limits = rev) +
    ylim(c(0, 100)) +
    geom_hline(yintercept=15, color = "grey", linewidth=2)
```

## tRNA/rRNA mapping rate

Samples should have a ribosomal RNA (rRNA) "contamination" rate below 10% (the grey line).

```{r plot_rrna_mapping_rate}

rrna_ylim <- max(round(metrics$r_and_t_rna_rate*100, 2)) + 10
metrics %>%
  ggplot(aes(x = factor(sample, level = order),
             y = r_and_t_rna_rate * 100, 
             color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  ylab("tRNA/rRNA rate, %")+
  ylim(0, rrna_ylim) + 
  ggtitle("tRNA/rRNA mapping rate") +
  coord_flip() +
  scale_x_discrete(limits = rev) +
  ylim(c(0, 100)) + xlab("") +
  geom_hline(yintercept=10, color = "grey", linewidth=2)
```

## 5'->3' bias

There should be little bias, i.e. the values should be close to 1, or at least consistent among samples

```{r plot_53_bias, eval=!all(is.na((metrics['r_and_t_rna_rate'])))}
metrics %>%
  ggplot(aes(x = factor(sample, level = order),
             y = x5_3_bias, 
             color = .data[[factor_of_interest]])) +
  geom_point(alpha = 0.5, size=4) +
  ggtitle("5'-3' bias") + 
  coord_flip() +
  scale_x_discrete(limits = rev) +
  ylim(c(0.5,1.5)) + xlab("") + ylab("5'-3' bias") +
  geom_hline(yintercept=1, color = "grey", linewidth=2)
```

## Counts per gene - all genes

We expect consistency in the box plots here between the samples, i.e. the distribution of counts across the genes is similar

```{r plot_counts_per_gene}
metrics_small <- metrics %>% dplyr::select(sample, .data[[factor_of_interest]])
metrics_small <- left_join(sample_names, metrics_small)

counts_lng <- 
  counts %>% 
  as_tibble() %>%
  filter(rowSums(.)!=0) %>% 
  gather(name, counts) 

counts_lng <- left_join(counts_lng, metrics_small, by = c("name" = "sample"))

ggplot(counts_lng, aes(factor(name, level = order),
                   log2(counts+1),
                   fill = .data[[factor_of_interest]])) +
  geom_boxplot() + 
  scale_x_discrete(limits = rev) +
  coord_flip() + xlab("") +
  ggtitle("Counts per gene, all non-zero genes")
```


# Sample similarity analysis

In this section, we look at how well the different groups in the dataset cluster with each other. Samples from the same group should ideally be clustering together. We use Principal Component Analysis (PCA).

## Principal component analysis (PCA) 

Principal Component Analysis (PCA) is a statistical technique used to simplify high-dimensional data by identifying patterns and reducing the number of variables. In the context of gene expression, PCA helps analyze large datasets containing information about the expression levels of thousands of genes across different samples (e.g., tissues, cells).

<!-- ### PCA, PCs 1-4, (labled) -->
```{r PCA1:5 summary, all, unlabeled, fig.width= 7, fig.height = 5}

vst <- vst(counts)

coldat_for_pca <- as.data.frame(metrics)
rownames(coldat_for_pca) <- coldat_for_pca$sample
coldat_for_pca <- coldat_for_pca[colnames(counts),]
pca1 <- degPCA(vst, coldat_for_pca,
              condition = factor_of_interest, data = T)[["plot"]]
pca2 <- degPCA(vst, coldat_for_pca,
              condition = factor_of_interest, data = T, pc1="PC3", pc2="PC4")[["plot"]]



pca1 + scale_color_grafify(palette = "kelly")
pca2 + scale_color_grafify(palette = "kelly")

```

# Covariates analysis

When there are multiple factors that can influence the results of a given experiment, it is useful to assess which of them is responsible for the most variance as determined by PCA. This method adapts the method described by Daily et al. for which they integrated a method to correlate covariates with principal components values to determine the importance of each factor. 

```{r covariate-plot,fig.height=12, fig.width=10}
## Remove non-useful columns output by nf-core
coldat_2 <- data.frame(coldat_for_pca[,!(colnames(coldat_for_pca) %in% c("fastq_1", "fastq_2", "salmon_library_types", "salmon_compatible_fragment_ratio", "samtools_reads_mapped_percent", "samtools_reads_properly_paired_percent", "samtools_mapped_passed_pct", "strandedness", "qualimap_5_3_bias"))])

# Remove missing data
coldat_2 <- na.omit(coldat_2)
degCovariates(vst, metadata = coldat_2)
```

## Hierarchical clustering

Inter-correlation analysis (ICA) is another way to look at how well samples
cluster by plotting the correlation between the expression profiles of the
samples.

```{r clustering fig, fig.width = 10, fig.asp = .62}
vst_cor <- cor(vst)

colma=coldata %>% as.data.frame()
rownames(colma) <- colma$sample
colma <- colma[rownames(vst_cor), ]
colma <- colma %>% dplyr::select(.data[[factor_of_interest]])
anno_colors=lapply(colnames(colma), function(c){
  l.col=grafify:::graf_palettes[["kelly"]][1:length(unique(colma[[c]]))]
  names(l.col)=unique(colma[[c]])
  l.col
})
names(anno_colors)=colnames(colma)

p <- pheatmap(vst_cor, 
         annotation = colma,
         annotation_colors = anno_colors,
         show_rownames = T, 
         show_colnames = T, 
         color = colorRampPalette(brewer.pal(11, "RdBu"))(15)
         )
p
```

# R session

List and version of tools used for the QC report generation.

```{r}
sessionInfo()
```
