---
title: "Weighted Gene Co-Expression Network Analysis (WGCNA)"
author: "Harvard Chan Bioinformatics Core"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
editor_options: 
  chunk_output_type: inline
params:
  minModuleSize: 30
  maxBlockSize: 4000
  mergeCutHeight: 0.25
  column: "sample_type"
  project_file: ../information.R
  params_file: params_de-example.R
  functions_file: ../libs 
---

```{r, message=FALSE, warning=FALSE}
# This set up the working directory to this file so all files can be found
library(rstudioapi)
library(tidyverse)
setwd(fs::path_dir(getSourceEditorContext()$path))
# NOTE: This code will check version, this is our recommendation, it may work
#.      other versions
stopifnot(R.version$major>= 4) # requires R4
if (compareVersion(R.version$minor,"3.1")<0) warning("We recommend >= R4.3.1") 
stopifnot(compareVersion(as.character(BiocManager::version()), "3.18")>=0)
```

This code is in this ![](https://img.shields.io/badge/status-alpha-yellow) revision.


```{r load_params, cache = FALSE, message = FALSE, warning=FALSE}
source(params$project_file)
source(params$params_file)
map(list.files(params$functions_file,pattern = "*.R$",full.names = T),source) %>% invisible()
column=params$column
contrasts=params$contrasts
subset_column=params$subset_column
subset_value=params$subset_value
```

```{r sanitize_datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
 DT::datatable(df, ..., rownames=gsub("-", "_", rownames(df)),
                   colnames=gsub("-", "_", colnames(df)))
}
```


```{r load_libraries, cache = FALSE, message = FALSE, warning=FALSE}
library(WGCNA)
library(tidyverse)
library(DESeq2)
library(magrittr)
library(rtracklayer)
library(flashClust)
library(gridExtra)
library(DT)
library(AnnotationDbi)
library(bcbioR)

library(knitr)
library(ggprism)


colors=cb_friendly_cols(1:15)
ggplot2::theme_set(theme_prism(base_size = 14))
opts_chunk[["set"]](
    cache = F,
    cache.lazy = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE,
    echo = T, 
    fig.height = 4)

# set seed for reproducibility
set.seed(1234567890L)
```

# Overview

-   Project: `r project`
-   PI: `r PI`
-   Analyst: `r analyst`
-   Experiment: `r experiment`
-   Aim: `r aim`

```{r load_data}

coldata <- load_coldata(coldata_fn, column,
                        subset_column, subset_value)
#coldata[[contrasts[[1]][1]]] = relevel(as.factor(coldata[[contrasts[[1]][1]]]), contrasts[[1]][3])
coldata$sample=row.names(coldata)

counts <- load_counts(counts_fn)
counts <- counts[,colnames(counts) %in% coldata$sample]

# It's easier for future items if the metadata(genotype) is already set up as a factor
#coldata$sample_type <- as.factor(coldata$sample_type)

```

# WGCNA

WGCNA carries out a correlation network construction. Networks are visual representations of interactions between `nodes` in a system. The nodes in WGCNA are individual genes. So, it helps to visualize patterns and relationships between gene expression profiles to identify genes associated with measured traits as well as identifying genes that are consistently co-expressed and could be contributing to similar molecular pathways. It also accounts for inter-individual variation in gene expression and alleviates issues associated with multiple testing.

We will use DESeq2 to transform our RNA-seq count data before running WGCNA. Removing the low count genes before the WGCNA is recommended and can help improve the WGCNA results. 

For this analysis we are keeping genes with **10 or more reads in total and expressed in at least 3 samples.**

For the data normalization we will use variance stabilizing transformation as recommended by the WGCNA's authors and transpose the dataset to prepare for WGCNA analysis. 


# Data

```{r show_coldata}
coldata %>% sanitize_datatable()
```

```{r normalize_data}
dds <- DESeqDataSetFromMatrix(counts, 
                              colData = coldata, 
                              design = ~ 1)

## Filtering lowly expressed genes
# We are filtering out genes with fewer than 10 raw counts in total and are present in fewer than 3 samples.
keep <- rowSums(counts(dds)>=10) >=4
dds <- dds[keep, ]

# Retrive the normalized data from the DESeqDataSet
dds_norm <- vst(dds)
normalized_counts <- assay(dds_norm) %>%
  t() # Transpose this data
#At this point you can look into the data and remove any outlier samples, as those outlier samples can affect the WGCNA results. However, our dataset as analysed earlier doesnot have obvious outliers.

```

## Parameter settings for WGCNA
WGCNA creates a weighted network to define which genes are near each other. The measure of adjacency used is based on the correlation matrix, which requires the definition of a threshold value, which in turn depends on a "power" parameter that defines the exponent used when transforming the correlation values. 

The choice of power parameter will affect the number of modules identified. WGCNA has a function called pickSoftThreshold() to help determine the soft power threshold.

```{r, message=FALSE, warning=FALSE,fig.align='center'}

# determine parameters for WGCNA
sft <- pickSoftThreshold(normalized_counts,
  dataIsExpr = TRUE,
  corFnc = cor,
  networkType = "signed"
)
# We have to calculate a measure of the model fit, the signed R2, and make that a new variable.

sft_df <- data.frame(sft$fitIndices) %>%
  dplyr::mutate(model_fit = -sign(slope) * SFT.R.sq)


ggplot(sft_df, aes(x = Power, y = model_fit, label = Power)) +
  geom_point() +
  geom_text(nudge_y = 0.1) +
  geom_hline(yintercept = 0.80, col = "red") +
  ylim(c(min(sft_df$model_fit), 1.1)) +
  xlab("Soft Threshold (power)") +
  ylab("Scale Free Topology Model Fit, signed R^2") +
  ggtitle("Scale independence") +
  theme_classic()
```

It is recommend to use a power that has an signed R2 above 0.80, otherwise the results may be too noisy to draw any meanings. In case of multiple power values with signed R2 above 0.80, we need to pick the one at an inflection point or where the R2 value seem to have reached their saturation. You want a power that gives a big enough R2 but is not excessively large. 

## One-step blockwise module detection:

We will use the blockwiseModules() with power threshold of 20 to find co-expressed genes. It will construct modules with group of genes that are highly correlated.

```{r, message=FALSE, warning=FALSE}
# picking a soft threshold power near the curve of the plot 
picked_power = 20  ## pick a power here based on the above plot and instruction
temp_cor <- cor
cor <- WGCNA::cor         
bwnet <- blockwiseModules(normalized_counts,                

                          # == Adjacency Function ==
                          power = picked_power,                
                          networkType = "signed", # there is option to run unsigned networktype as well.

                          # == Set Tree and Block parameters ==
                          deepSplit = 2,
                          pamRespectsDendro = F,
                          # detectCutHeight = 0.75,
                          minModuleSize = params$minModuleSize,
                          maxBlockSize = params$maxBlockSize,

                          # == Module Adjustments ==
                          reassignThreshold = 0,
                          mergeCutHeight = params$mergeCutHeight,

                          # == TOM == Archive the run results in TOM file (saves time)
                        # saveTOMs = T,
                        # saveTOMFileBase = "ER",

                          # == Output Options
                          numericLabels = T,
                          verbose = F)

# Write down main WGCNA results object to file
readr::write_rds(bwnet,
                 file = "WGCNA_results.RDS")
```

## Eigengene module dataframe

A module eigengene is the standardized gene expression profile for a given module. An eigengene is the gene whose expression is representative of the majority of the genes within a module.

```{r, message=FALSE, warning=FALSE}
module_eigengens <- bwnet$MEs
datatable(module_eigengens, options = list(pageLength =10, scrollX=TRUE))

```

## Dendogram for the modules

```{r, fig.align='center', fig.width=12, message=FALSE, warning=FALSE}
# Convert labels to colors for plotting
mergedColors =labels2colors(bwnet$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(bwnet$dendrograms[[1]],
                    mergedColors[bwnet$blockGenes[[1]]],
                    "Module colors",
                    dendroLabels = FALSE,
                    hang = 0.03,
                    addGuide = TRUE,
                    guideHang = 0.05)

```

### Relating modules (Clusters) to geneIds. 
GeneIds with their module color and number are given on the table below.

```{r, message=FALSE, warning=FALSE}
# Relating modules (Clusters) to genotypes
module_df <- data.frame(
  gene_id = names(bwnet$colors),
  module_no = bwnet$colors,
  colors = labels2colors(bwnet$colors)
  
)
datatable(module_df, rownames = FALSE)
# lets write out the file lising the genes and their modules
write_delim(module_df, file = "List_of_genes_with_modules.txt", delim = "\t")
```

## Relating modules with genotypes

WGCNA calcualtes and Eigengene (hypothetical central gene) for each module which makes it easier to associate sample groups with module cluster.

```{r, message=FALSE, warning=FALSE}
# Get Module Eigengens per cluster
MEs0 <- moduleEigengenes(normalized_counts, mergedColors)$eigengenes

# Reorder modules so similar modules are next to each other
MEs0 <- orderMEs(MEs0)
module_order = names(MEs0) %>% gsub("ME","", .)

# Add sample names
MEs0$samples = row.names(MEs0)

# tidy the data
mME = MEs0 %>%
  pivot_longer(-samples) %>%
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )
```

```{r, fig.align='center', fig.height=8, message=FALSE, warning=FALSE}
mME %>% ggplot(., aes(x=samples, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red3",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")

```

On the heatmap above, samples are on the x-axis and different modules on y-axis.
We can visually inspect this heatmap to understand how different modules are correlated to different sample groups. 

We can also use another approach to extract only the modules with biggest differences and then correlate those modules for conditions.

## Modules with biggest differences across treatments
We will create a design matrix using `sample_type` in our metadata and run linear model on each module.

```{r, message=FALSE, warning=FALSE}
# to confirm our eigengens relate to our metadata from deseq object run the line below
#all.equal(colnames(dds), rownames(module_eigengens))

# Create a design matrix from the genotype variable
des_mat <- model.matrix(~ coldata$sample_type)


# lmFit() needs a transposed version of the matrix
fit <- limma::lmFit(t(module_eigengens), design = des_mat)

# Apply empirical Bayes to smooth standard errors
fit <- limma::eBayes(fit)

# Apply multiple testing correction and obtain stats
stats_df <- limma::topTable(fit, number = ncol(module_eigengens)) %>%
  tibble::rownames_to_column("module")

datatable(stats_df, options = list(pageLength =10, scrollX=TRUE))
```

Modules with highest padj values are the ones have bigger differences between sample_groups. Look for these modules in the heatmap above.

# Conclusions



# Methods

RNA-seq counts were generated by the nf-core rnaseq pipeline [version] using Salmon (Patro et al. 2017). Downstream analyses were performed using `r version$version.string`. Counts were imported into R using DESeq2 version `r packageVersion("DESeq2")` (Love, Huber, and Anders 2014). Gene annotations were obtained from Ensembl. Plots were generated by ggplot2 (Wickham 2016). Heatmaps were generated by pheatmap (Kolde 2019).

## R package references

```{r citations}
citation("WGCNA")
citation("DESeq2")
citation("ggplot2")
```

## R session

List and version of tools used for the QC report generation.

```{r}
sessionInfo()
```
