---
title: "Immune Deconvolution"
author: "Harvard Chan Bioinformatics Core"
date: "`r Sys.Date()`"
output:
   html_document:
      code_folding: hide
      df_print: paged
      highlights: pygments
      number_sections: true
      self_contained: true
      theme: default
      toc: true
      toc_float:
         collapsed: true
         smooth_scroll: true
params:
  # information about project: project name, PI, analyst, experiment, aim
  project_file: ../information.R
  # paths to raw counts and TPM gene expression matrices and metadata table
  params_file: params_immune_deconv.R
  # metadata to separate by on plot
  factor_of_interest_1: tissue
  factor_of_interest_1_levels: c("Duo", "Ile")
  factor_of_interest_2: timepoint
  factor_of_interest_2_levels: c("0d", "2d", "4d", "6d", "8d", "2w", "3w", "4w")
  # which organism are your expression matrices from?
  # can be human or mouse
  organism: mouse
  # for TIMER method: which cancer type?
  # run `immunedeconv::timer_available_cancers` to see options
  # or see publication https://doi.org/10.1186/s13059-016-1028-7 for definitions
  cancer_type: coad
---

Template developed with materials from the [Immune deconvolution vignette from immunedeconv](https://omnideconv.org/immunedeconv/articles/immunedeconv.html) and the [Immune infiltration tutorial from DFCI](https://liulab-dfci.github.io/RIMA/Infiltration.html).

```{r check_wd_and_version}
# this sets up the working directory to this file so all files can be found
library(rstudioapi)
setwd(fs::path_dir(getSourceEditorContext()$path))
# NOTE: this code will check version, this is our recommendation, it may work with other versions
stopifnot(R.version$major >= 4) # requires R4
if (compareVersion(R.version$minor, "3.1") < 0) warning("We recommend >= R4.3.1") 
stopifnot(compareVersion(as.character(BiocManager::version()), "3.18") >= 0)
```

This code is in this ![](https://img.shields.io/badge/status-draft-grey) revision.

```{r load_params, cache = FALSE}
# 1. set up input files in this R file (line 20: params_immune_deconv.R)
source(params$params_file)
# 2. set up project file (already done from QC probably)
source(params$project_file)
# IMPORTANT: set these values if you are not using the parameters in the header (lines 20-35)
# input_counts <- "../counts/tximport-counts.csv"
# input_tpm    <- "../counts/tximport-tpm.csv"
# metadata <- "metadata.csv"
factor_of_interest_1        <- params$factor_of_interest_1
factor_of_interest_1_levels <- params$factor_of_interest_1_levels
factor_of_interest_2        <- params$factor_of_interest_2
factor_of_interest_2_levels <- params$factor_of_interest_2_levels
organism    <- params$organism
cancer_type <- params$cancer_type

# determine which gene conversion chunks to run based on starting organism
run_mouse <- if(organism == "mouse") TRUE else FALSE
run_human <- if(organism == "human") TRUE else FALSE
```

```{r load_libraries, cache = FALSE, message = FALSE, warning=FALSE}
library(tidyverse)
# BiocManager::install('grimbough/biomaRt')
library(biomaRt)
# remotes::install_github("omnideconv/immunedeconv")
library(immunedeconv)

colors <- cb_friendly_cols(1:15)
ggplot2::theme_set(theme_prism(base_size = 14))
opts_chunk[["set"]](
    cache = FALSE,
    cache.lazy = FALSE,
    dev = c("png", "pdf"),
    error = TRUE,
    highlight = TRUE,
    message = FALSE,
    prompt = FALSE,
    tidy = FALSE,
    warning = FALSE,
    echo = TRUE, 
    fig.height = 4)

# set seed for reproducibility
set.seed(1234567890L)
```

```{r sanitize_datatable}
sanitize_datatable = function(df, ...) {
 # remove dashes which cause wrapping
  DT::datatable(df, ...,
                rownames = gsub("-", "_", rownames(df)),
                colnames = gsub("-", "_", colnames(df)))
}
```

# Overview

-   Project: `r project`
-   PI: `r PI`
-   Analyst: `r analyst`
-   Experiment: `r experiment`
-   Aim: `r aim`

```{r load_metadata}
# load metadata
metadata <- read.csv(metadata)

# make metadata columns into factors (so samples plot in the right order)
metadata[, factor_of_interest_1] <- factor(metadata[, factor_of_interest_1],
                                           levels = factor_of_interest_1_levels)
metadata[, factor_of_interest_2] <- factor(metadata[, factor_of_interest_2],
                                           levels = factor_of_interest_2_levels)
```

```{r load_gene_expression}
# load gene expression matrix as data frame
df_counts <- read.table(input_counts, sep = ",", row.names = 1, header = TRUE, check.names = FALSE)
df_tpm    <- read.table(input_tpm,    sep = ",", row.names = 1, header = TRUE, check.names = FALSE)
```

```{r convert_gene_names_mouse, eval = run_mouse}
# immunedeconv methods expect an input of a gene expression matrix as a data.frame or matrix where:
#   rownames = MGI (for mouse methods) or HGNC (for human methods) symbols
#   colnames = sample names
# we currently have mouse Ensembl IDs as rownames
# so will need to convert to MGI

# get mouse reference
ensembl_mouse <- useEnsembl(biomart = "genes", dataset = "mmusculus_gene_ensembl")
genenames_mouse <- getBM(values     = rownames(df_counts), # vector of gene IDs (same for "counts" and "tpm")
                         filters    = "ensembl_gene_id",   # type of gene ID we are using
                         attributes = c("ensembl_gene_id", # attributes we want to retrieve
                                        "mgi_symbol"),
                         mart       = ensembl_mouse)       # mart we are using (defined above)

# convert from mouse Ensembl ID (ENMUSG) to MGI ID (gene symbol)
convert_gene_names_mouse_ensembl_to_MGI <- function(df) {
  # convert gene names to MGI
  df_clean <- df %>% rownames_to_column(var = "ensembl_gene_id") %>%
    # remove existing rownames
    as_tibble() %>%
    # add MGI gene symbols
    left_join(genenames_mouse, by = "ensembl_gene_id") %>%
    # remove Ensembl gene IDs
    dplyr::select(-ensembl_gene_id) %>%
    # keep only distinct rows
    distinct() %>%
    # drop any rows where we now don't have gene symbols
    subset(!is.na(mgi_symbol) & mgi_symbol != "")
  
  # combine any rows where multiple mouse Ensembl IDs map to the same gene symbol
  df_clean <- df_clean %>% group_by(mgi_symbol) %>%
    summarize(across(starts_with(c("Duo", "Ile")), sum))
  
  # turn MGI symbols back into rownames
  df_clean <- df_clean %>% column_to_rownames("mgi_symbol")
  
  return(df_clean)
}

# convert from mouse Ensembl ID to mouse MGI symbol
df_counts_mouse <- convert_gene_names_mouse_ensembl_to_MGI(df_counts)
df_tpm_mouse    <- convert_gene_names_mouse_ensembl_to_MGI(df_tpm)

# also convert from mouse MGI symbol to human HGNC symbol so we can run additional methods
df_counts_human <- convert_human_mouse_genes(df_counts_mouse, convert_to = "human")
df_tpm_human    <- convert_human_mouse_genes(df_tpm_mouse,    convert_to = "human")
```

```{r convert_gene_names_human, eval = run_human}
# immunedeconv methods expect an input of a gene expression matrix as a data.frame or matrix where:
#   rownames = MGI (for mouse methods) or HGNC (for human methods) symbols
#   colnames = sample names
# we currently have human Ensembl IDs as rownames
# so will need to convert to HGNC

# get human reference
ensembl_human <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
genenames_human <- getBM(values     = rownames(df_counts), # vector of gene IDs (same for "counts" and "tpm")
                         filters    = "ensembl_gene_id",   # type of gene ID we are using
                         attributes = c("ensembl_gene_id", # attributes we want to retrieve
                                        "hgnc_symbol"),
                         mart       = ensembl_human)       # mart we are using (defined above)

# convert from human Ensembl ID (ENSG) to HGNC ID (gene symbol)
convert_gene_names_human_ensembl_to_HGNC <- function(df) {
  # convert gene names to HGNC
  df_clean <- df %>% rownames_to_column(var = "ensembl_gene_id") %>%
    # remove existing rownames
    as_tibble() %>%
    # add HGNC gene symbols
    left_join(genenames_human, by = "ensembl_gene_id") %>%
    # remove Ensembl gene IDs
    dplyr::select(-ensembl_gene_id) %>%
    # keep only distinct rows
    distinct() %>%
    # drop any rows where we now don't have gene symbols
    subset(!is.na(hgnc_symbol) & hgnc_symbol != "")
  
  # combine any rows where multiple human Ensembl IDs map to the same gene symbol
  df_clean <- df_clean %>% group_by(hgnc_symbol) %>%
    summarize(across(starts_with(c("Duo", "Ile")), sum))
  
  # turn HGNC symbols back into rownames
  df_clean <- df_clean %>% column_to_rownames("hgnc_symbol")
  
  return(df_clean)
}

# convert from human Ensembl ID to human HGNC symbol
df_counts_human <- convert_gene_names_human_ensembl_to_HGNC(df_counts)
df_tpm_human    <- convert_gene_names_human_ensembl_to_HGNC(df_tpm)

# also convert from human HGNC symbol to mouse MGI symbol so we can run additional methods
df_counts_mouse <- convert_human_mouse_genes(df_counts_human, convert_to = "mouse")
df_tpm_mouse    <- convert_human_mouse_genes(df_tpm_human,    convert_to = "mouse")
```

```{r function_graph_results}
graph_deconvolution_results <- function(res, method){
  # reshape data to one row per sample per cell type
  res_long <- res %>%
    gather(sample, score, -cell_type) %>%
    # add metadata columns
    left_join(metadata, by = "sample")
  
  # graph
  print(
    res_long %>%
      ggplot(aes(x = .data[[factor_of_interest_2]], y = score,
                 color = cell_type, shape = .data[[factor_of_interest_1]])) +
      geom_point(size = 4) +
      ggtitle(paste0(method, ": Immune cell deconvolution")) +
      facet_wrap(~cell_type, scales = "free", ncol = 3) +
      scale_color_discrete(guide = "none") +
      theme_bw() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
  )
  
  # also graph subsets by factor_of_interest_1 (e.g. tissue type)
  for(i in 1:length(factor_of_interest_1_levels)) {
    print(
      res_long %>% subset(.data[[factor_of_interest_1]] == factor_of_interest_1_levels[i]) %>%
        ggplot(aes(x = .data[[factor_of_interest_2]], y = score, color = cell_type)) +
        geom_point(size = 4) +
        ggtitle(paste0(method, ": Immune cell deconvolution in ", factor_of_interest_1_levels[i])) +
        facet_wrap(~cell_type, scales = "free", ncol = 3) +
        scale_color_discrete(guide = "none") +
        theme_bw() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
    )
  }
}
```

# Immune deconvolution using mouse genes

When using mouse MGI gene symbols, we have a total of `r nrow(df_counts_mouse)` genes available to test. `r if(organism == "human") cat("The number here is much smaller because not all human genes can be mapped to mouse genes.")`

The following methods are currently available for mouse data: `r paste(names(deconvolution_methods_mouse), collapse = ", ")`.

## mMCPcounter

mMCPcounter is a mouse data-based method (related to the human method MCPcounter) using **`r nrow(df_counts_mouse)` genes**.

mMCPcounter provides cell-type scores for each sample in arbitrary units that are **comparable between samples** but not between cell-types.

```{r mmcp_counter, fig.width = 10, fig.height = 10}
# TPM normalized is preferred input
res_mmcp <- immunedeconv::deconvolute_mouse(df_tpm_mouse, "mmcp_counter")

res_mmcp %>% sanitize_datatable()

graph_deconvolution_results(res_mmcp, "mMCPcounter")

# save results
write.csv(res_mmcp, file = "results/mouse_mMCPcounter.csv", quote = FALSE)
```

## SeqImmuCC

SeqImmuCC is a mouse data-based method using **`r nrow(df_counts_mouse)` genes**.

SeqImmuCC provides cell-type scores for each sample in units that are **comparable between both samples and cell-types**.

```{r seqimmucc, fig.width = 10, fig.height = 8}
# raw counts is preferred input
# requires additional parameter "algorithm": 'SVR' for CIBERSORT or 'LLSR' for least squares regression
res_seqimmucc <- immunedeconv::deconvolute_mouse(df_counts_mouse, "seqimmucc", algorithm = "LLSR")

res_seqimmucc %>% sanitize_datatable()

graph_deconvolution_results(res_seqimmucc, "SeqImmuCC")

# save results
write.csv(res_seqimmucc, file = "results/mouse_SeqImmuCC.csv", quote = FALSE)
```

## DCQ

DCQ is a mouse data-based method using **`r nrow(df_counts_mouse)` genes**.

DCQ provides cell-type scores for each sample in arbitrary units that are **comparable between cell-types** but not between samples.

```{r dcq, fig.width = 10, fig.height = 12}
# TPM normalized is preferred input
res_dcq <- immunedeconv::deconvolute_mouse(df_tpm_mouse, "dcq")

res_dcq %>% sanitize_datatable()

graph_deconvolution_results(res_dcq, "DCQ")

# save results
write.csv(res_dcq, file = "results/mouse_DCQ.csv", quote = FALSE)
```

## BASE

BASE is a mouse data-based method using **`r nrow(df_counts_mouse)` genes**.

BASE provides cell-type scores for each sample in arbitrary units that are **comparable between samples** but not between cell-types.

```{r base, fig.width = 10, fig.height = 10}
# TPM normalized is preferred input
res_base <- immunedeconv::deconvolute_mouse(df_tpm_mouse, "base")

res_base %>% sanitize_datatable()

graph_deconvolution_results(res_base, "BASE")

# save results
write.csv(res_base, file = "results/mouse_BASE.csv", quote = FALSE)
```

# Immune deconvolution using human genes

When using human HGNC gene symbols, we have a total of `r nrow(df_counts_human)` genes available to test. `r if(organism == "mouse") cat("The number here is much smaller because not all mouse genes can be mapped to human genes.")`

The following methods are currently available for human data: `r paste(names(deconvolution_methods), collapse = ", ")`. `r if(organism == "mouse") cat("The reason to convert to human genes is that more methods are available for immune deconvolution.")`

## CIBERSORT

CIBERSORT is a human data-based method using **`r nrow(df_counts_human)` genes**.

Unlike all the other modules, which were run in R in this report, CIBERSORT was run using [the online portal](https://cibersortx.stanford.edu/runcibersortx.php) due to its unique licensing agreement. The following parameters were used:

- Job type: Impute Cell Fractions
- Signature matrix file: LM22.update-gene-symbols.txt (22 immune cell types, built-in reference)
- Mixture file: tpm_human.tsv
- Batch correction: disabled
- Disable quantile normalization: true
  - *Disabling is recommended for RNA-Seq data*
- Run mode (relative or absolute): absolute
  - *Description: "Absolute mode scales relative cellular fractions into a score that reflects the absolute proportion of each cell type in a mixture. Although not currently expressed as a fraction, the absolute score can be directly compared across cell types (i.e., relative differences between cell types are maintained) and the default output (relative fractions) can be easily re-derived by normalizing the absolute scores to 1."*
- Permutations: 100

CIBERSORT-abs provides cell-type scores for each sample in units that are **comparable between both samples and cell-types**.

```{r cibersort, fig.width = 10, fig.height = 16}
# load results from web portal
res_cibersort <- read.csv("results/human_CIBERSORT_web.csv", header = TRUE)

# these results have columns = celltypes and rows = samples, which is reversed
res_cibersort <- res_cibersort %>% column_to_rownames("Mixture") %>%
  t() %>% as.data.frame() %>% rownames_to_column(var = "cell_type")

# remove extra data that is not scores per sample / cell type
res_cibersort <- subset(res_cibersort, !(cell_type %in% c("P.value",
                                                          "Correlation",
                                                          "RMSE",
                                                          "Absolute.score..sig.score.")))

# clean up cell type names
res_cibersort$cell_type <- gsub("\\.", " ", res_cibersort$cell_type) # replace . with space
res_cibersort$cell_type <- gsub("  ", " ", res_cibersort$cell_type)  # replace two spaces with single space

res_cibersort %>% sanitize_datatable()

graph_deconvolution_results(res_cibersort, "CIBERSORT-abs")
```

## TIMER (Tumor IMmune Estimation Resource)

TIMER is a human data-based method using **`r nrow(df_counts_human)` genes**.

TIMER provides cell-type scores for each sample in arbitrary units that are **comparable between samples** but not between cell-types.

```{r timer, fig.width = 10, fig.height = 4}
# TPM normalized is preferred input
# requires additional parameter "indications": cancer type
res_timer <- immunedeconv::deconvolute(df_tpm_human, "timer",
                                       indications = rep(tolower(cancer_type), ncol(df_tpm_human)))

res_timer %>% sanitize_datatable()

graph_deconvolution_results(res_timer, "TIMER")

# save results
write.csv(res_timer, file = "results/human_TIMER.csv", quote = FALSE)
```

## quanTIseq

quanTIseq is a human data-based method using **`r nrow(df_counts_human)` genes**.

quanTIseq provides cell-type scores for each sample in absolute units that are **comparable between both samples and cell-types** and **can be interpreted as a cell fraction**.

```{r quantiseq, fig.width = 10, fig.height = 8}
# TPM normalized is preferred input
res_quant <- immunedeconv::deconvolute(df_tpm_human, "quantiseq",
                                       tumor = FALSE, arrays = FALSE)

res_quant %>% sanitize_datatable()

graph_deconvolution_results(res_quant, "quanTIseq")

# save results
write.csv(res_quant, file = "results/human_quanTIseq.csv", quote = FALSE)
```

## EPIC

EPIC is a human data-based method using **`r nrow(df_counts_human)` genes**.

EPIC provides cell-type scores for each sample in absolute units that are **comparable between both samples and cell-types** and **can be interpreted as a cell fraction**.

```{r epic, fig.width = 10, fig.height = 6}
# TPM normalized is preferred input
# requires additional parameter "tumor": tumor data T/F
res_epic <- immunedeconv::deconvolute(df_tpm_human, "epic",
                                      tumor = FALSE)

res_epic %>% sanitize_datatable()

graph_deconvolution_results(res_epic, "EPIC")

# save results
write.csv(res_epic, file = "results/human_EPIC.csv", quote = FALSE)
```

## MCPcounter

MCPcounter is a human data-based method (related to the mouse method mMCPcounter) using **`r nrow(df_counts_human)` genes**.

MCPcounter provides cell-type scores for each sample in arbitrary units that are **comparable between samples** but not between cell-types.

```{r mcp_counter, fig.width = 10, fig.height = 8}
# TPM normalized is preferred input
res_mcp <- immunedeconv::deconvolute(df_tpm_human, "mcp_counter")

res_mcp %>% sanitize_datatable()

graph_deconvolution_results(res_mcp, "MCPcounter")

# save results
write.csv(res_mcp, file = "results/human_MCPcounter.csv", quote = FALSE)
```

# Analysis using a custom gene set

Since MCPcounter (human) and mMCPcounter (mouse) are marker-based, we can provide a custom set of genes to perform deconvolution of cell types not included by the above algorithms.

As a reminder, these methods provide cell-type scores for each sample in arbitrary units that are **comparable between samples** but not between cell-types.

## Human data

For **human** data, the input gene list requires at least two columns (see [this example](https://raw.githubusercontent.com/ebecht/MCPcounter/master/Signatures/genes.txt) from the MCPcounter GitHub):

1. **Cell population**, which defines which cell type each gene is a marker for,

and one or more of

2a. **HUGO symbols**, the HGNC gene identifier (which should already match the rownames of `df_tpm_human`)
2b. **ENTREZID**
2c. **ENSEMBL ID**

```{r custom human}
# what does their built-in signature look like?
# https://raw.githubusercontent.com/ebecht/MCPcounter/master/Signatures/genes.txt

# load markers
# markers <- read.table("path/to/markers.txt", sep = "\t")
# we are using here as an example marker genes for innate lymphoid cells (ILCs)
markers <- data.frame(`Cell population` = rep(c("ILC1", "ILC2", "ILC3"), times = c(5, 7, 6)),
                      `HUGO symbols` = c("IL12", "IL15", "IL18", "IFNG", "TNF",                # ILC1
                                         "IL25", "IL33", "TSLP", "IL4", "IL5", "IL13", "AREG", # ILC2
                                         "IL1B", "IL23", "IL17", "IL22", "LTA", "CSF2"))       # ILC3

# check which marker genes are found in the human set
# df_tpm_human[which(rownames(df_tpm_human) %in% markers$`HUGO symbols`),]

# TPM normalized is preferred input
res_mcp_custom <- immunedeconv::deconvolute_mcp_counter(df_tpm_human,
                                                        # add our custom gene set
                                                        genes = markers,
                                                        # specify which gene identifiers we are using
                                                        # options: "HUGO_symbols", "ENTREZ_ID", "ENSEMBL_ID"
                                                        feature_types = "HUGO_symbols") %>%
  as.data.frame() %>% rownames_to_column("cell_type")

res_mcp_custom %>% DT::datatable()

graph_deconvolution_results(res_mcp_custom, "MCPcounter for custom gene set")

# save results
write.csv(res_mcp_custom, file = "results/human_MCPcounter_custom.csv", quote = FALSE)
```

## Mouse data

For **mouse** data, the input gene list requires at least two columns:

1. **Denomination**, which defines which cell type each gene is a marker for,

and one or more of

2a. **Gene.Symbol**, the MGI gene identifier (which should already match the rownames of `df_tpm_mouse`)
2b. **ENSEMBL.ID**

```{r custom mouse}
# what does their built-in signature look like?
# data("mMCPcounter_signatures_GCRm39", envir = sys.frame(sys.nframe()), package = "mMCPcounter")

# load markers
# markers <- read.table("path/to/markers.txt", sep = "\t")
# we are using here as an example marker genes for innate lymphoid cells (ILCs)
markers <- data.frame(Denomination = rep(c("ILC1", "ILC2", "ILC3"),
                                         times = c(6, 7, 7)),
                      Gene.Symbol = c("Il12a", "Il12b", "Il15", "Il18", "Ifng", "Tnf",           # ILC1
                                      "Il25", "Il33", "Tslp", "Il4", "Il5", "Il13", "Areg",      # ILC2
                                      "Il1b", "Il23a", "Il17a", "Il22", "Il22b", "Lta", "Csf2")) # ILC3

# check which marker genes are found in the mouse set
# df_tpm_mouse[which(rownames(df_tpm_mouse) %in% markers$Gene.Symbol),]

# specify which gene identifiers we are using
# options: "Gene.Symbol", "ENSEMBL.ID"
features <- "Gene.Symbol"

# run mMCPcounter by hand
# code adapted from: https://github.com/cit-bioinfo/mMCP-counter/blob/master/R/mMCP-counter.R
median.na <- function(x) { return(median(x, na.rm = TRUE)) }
cit.dfAggregate <- function(data, partition, MARGIN = 1, fAggreg = median.na) {
  cMARGIN <- setdiff(c(1, 2), MARGIN)
  n <- length(partition)
  N <- dim(data)[MARGIN]
  p <- dim(data)[cMARGIN]
  if (n != N)
    stop("ERROR - cit.dfAggregate : size of partition doesn't correspond to data dimension")
  l <- split(1:N, partition)
  d <- data
  if (MARGIN == 2) { d <- t(data) }
  d <- matrix(sapply(l,
                     function(i) if(length(i) == 1) { unlist(d[i, ]) } else { apply(d[i, ], 2, fAggreg) }),
              ncol = p, byrow = TRUE)
  d <- as.data.frame(d)
  rownames(d) <- names(l)
  names(d) <- dimnames(data)[[cMARGIN]]
  if (MARGIN == 2) { d <- as.data.frame(t(d)) }
  d
}
foundGenes <- intersect(markers[,features], rownames(df_tpm_mouse))
if(length(foundGenes) == 0){
  stop("No signature found in input row names. Please ensure the features are accurately set.")
}
absentSignatures <- setdiff(unique(markers$Denomination),
                            unique(markers[markers[,features] %in% rownames(df_tpm_mouse),
                                           "Denomination"]))
if(length(absentSignatures) > 0){
  warning(paste("No genes were found for population(s): ", paste(absentSignatures, collapse = ", "),
                ". Please ensure the genomeVersion argument is correct.", sep = ""))
}
localSig <- markers[markers[, features] %in% foundGenes, ]
res_mmcp_custom <- df_tpm_mouse[localSig[, features], ]
res_mmcp_custom <- cit.dfAggregate(res_mmcp_custom,
                                   localSig$Denomination,
                                   fAggreg = median.na)
res_mmcp_custom <- res_mmcp_custom[apply(res_mmcp_custom, 1,
                                         function(x){ sum(is.na(x)) }) < ncol(res_mmcp_custom),]

res_mmcp_custom <- res_mmcp_custom %>% as.data.frame() %>% rownames_to_column("cell_type")

res_mmcp_custom %>% DT::datatable()

graph_deconvolution_results(res_mmcp_custom, "mMCPcounter for custom gene set")

# save results
write.csv(res_mmcp_custom, file = "results/mouse_mMCPcounter_custom.csv", quote = FALSE)
```

# R session

List and version of tools used for the report generation.

```{r}
sessionInfo()
```
